---
title: "Lab 1"
author: "Lydia Bleifuss"
date: "1/9/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(here) ###tell camila about this
library(janitor)
library(kableExtra)
```

Read in noaa_fisheries.csv data

```{r}
##ALWAYS use read_csv NOT read.csv

###Use the head package becuse it doens't matter what file pathway, just look here ----> 

us_landings <- read_csv(here("data","noaa_fisheries.csv"))
```

Now...make a tidy version of the data frame:

```{r}
###Good to keep one tidy version of your data and then when drawing different things out, call them new data frames so we always have the nice version 

landings_tidy <- us_landings %>% 
  janitor::clean_names() %>% 
  #mutate can be used to add an aditional column with something new, but can also be used to reformate an existing colum 
  mutate(state = str_to_lower(state),
         afs_name = str_to_lower(afs_name)) %>% 
  #this is saying mutate the existing colum, changing upper characters to lower) and then updating names column in same way 
  mutate(dollars_num = parse_number(dollars_usd)) ###in Read R package, there are different things for parsing, which can pull numbers out of character columns, etc. etc. (parse_number)
```

Now let's just get some informaiton for salmon: 

```{r}
salmon_landings <- landings_tidy %>% 
   #now, going to get rid of any entries that aren't specific to species level (basically if it says "aggregate" in the column afs_name, take out that word from that row) 
  mutate(afs_clean = str_remove(afs_name, pattern="aggregate")) %>% 
  #now I want to use filter to only keep the rows that I am interested in, asking, is this string detected, the string being "salmon" in this case 
  filter(str_detect(afs_clean, pattern="salmon")) %>% 
  #now I want to seperate the salmon from the specific species
  separate(afs_clean, into = c("group","species"), sep=",")
```


Find some group summary data: 

now we want to do some summarising (in excel, it's called pivot tables but here is is different)

Find annual total US landings and dollar value (summing across al states) for each TYPE of salmon using 'group-by()' + 'summarize()'

```{r}
#if I want to find the annual total across states for each salmon species then I need to group by salmon species 

salmon_summary <- salmon_landings %>% 
  group_by(year,species) %>% 
  #this doesn't look any different, we are just creating invisible group in R's brain, so when we use summarize() it uses those groups
  summarize(
    tot_landings = sum(landings_pounds), 
    tot_value = sum(dollars_num)
  )
```


Make a graph!

```{r}
salmon_landings_graph <- ggplot(data = salmon_summary, 
                                aes(x = year, y = tot_landings)) +
  geom_line(aes(color = species))
#need to tell we are using ggplot, what data we are using and what type of graph
#here, we also needed to say that we grouped different species, and we said that by group = x right after x and y in aes) -> can also do this in geom _line(aes(color = species)), either one don't need it in both places

salmon_landings_graph
```


Now we can export our graph, it will aplly to the most recently run gg plot code, or we could name our graph and export it specifically 

```{r}
ggsave(plot = salmon_landings_graph,
       here("figures", "us_salmon_LB.jpg"),
       height =5,
       width = 8) #we can also say explicitly where to send the graph we are exporting, perhaps the "figures" folder that came along when we cloned the repo 
```

We can also make a nice kable table: 

```{r}
#if I only want to head the first top lines in a function, we can do "head(30)" to keep first 30 and "tai(10)" to keep last 10 lines

salmon_first_5 <- salmon_summary %>% 
  head(5)
```

Use kable:

```{r}
kable(salmon_first_5) %>% 
  kable_styling(bootstrap_options = "striped",
                full_width =FALSE)
```
